---
title: "Ichthyosaur disparity"
author: [Ben Moon]
institute: [University of Bristol]
lang: english
tags: [Ichthyosauriformes, macroevolution]

titlepage: true
titlepage-color: 16a085
titlepage-text-color: ffffff
titlepage-rule-color: ffffff
titlepage-rule-height: 1

tablenos-cleveref: On
xnos-cleveref-fake: Off

bibliography: Bibliography.bib
csl: palaeontology.csl

abstract: "Using a cladistic matrix of ichthyosaurs, I analyse the disparity of ichthyosaurs throught the Mesozoic by use of morphospace and disparity."

output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
    theme: journal
    highlight: zenburn
  md_document:
    variant: markdown
    preserve_yaml: TRUE
    toc: TRUE
    md_extension: NULL
---

The document can be compiled into an HTML notebook using the following command in R.

```{r compile, eval = FALSE}
rmarkdown::render("ichthyosaur-disparity.Rmd", encoding = "utf8", output_format = "html_notebook")
```

First, install the required libraries. For this, we need `Claddis` and `dispRity`, which can both be downloaded from their respective GitHub repositories. This also requires `devtools`. If you want to parallelise the code, then adding in the package `parallel` is needed.

```{r install-libraries, eval = FALSE}
if(!require(devtools)) {install.packages("devtools")
	library(devtools)
	install_github("graemetlloyd/Claddis")
	install_github("TGuillerme/dispRity", ref = "release")
}
```

Now I load the required libraries.

```{r load-libraries, collapse = TRUE}
library(Claddis)
library(dispRity)
library(parallel)
library(knitr)
library(pander)
library(stringr)

## Calculate number of cores to use (1 less that available)
no_cores <- detectCores() - 1

## Settings for a code chunks
opts_chunk$set(tidy = TRUE, autodep = TRUE, dpi = 300)

getwd()
```

```{r child = "rmd/ana1-ordinate-data.Rmd"}
```

```{r child = "rmd/ana2-morphospace-xy.Rmd"}
```

```{r child = "rmd/ana3-time-binning.Rmd"}
```

```{r child = "rmd/ana4-disparity-time-sumvar.Rmd"}
```

```{r child = "rmd/ana5-disparity-time-centroid.Rmd"}
```

```{r child = "rmd/ana6-disparity-time-diagonal.Rmd"}
```

```{r child = "rmd/ana7-disparity-tests.Rmd"}
```

# References