# Disparity analyses

## Sum of variances

Then I get to the disparity analyses themselves. Firstly, I've used sum of variances as this metric has little variation for the bin population (i.e. is constant in rarefaction). These analyses can be quite time consuming, so I've parallelised them to reduce that somewhat. The avoid parallel tasks, comment out the lines with cluster functions, replace `parLapply` with `lapply`, and remove the first argument on each  `lapply`. For example, `parLapply(cl, x, y)` becomes `lapply(x, y)`.

```{r sumvar-disparity-analyses, cache = TRUE}
## Set up function to calculate disparity in parallel
parDispRitySV <- function(x) {
    dispRity(x, metric = c(sum, variances))
}

## For parallelisation: create cluster
clusDisparity <- makeCluster(no_cores)
## Export required variables
clusterExport(clusDisparity, c("dispRity", "variances"))

## Calculate disparity in epochs
disparitySV_epoch <- parLapply(clusDisparity, bootstrapped_epoch, parDispRitySV)
## Calculate disparity fro bJ binning
disparitySV_bJ <- parLapply(clusDisparity, bootstrapped_bJ, parDispRitySV)

## Release cluster cores
stopCluster(clusDisparity)
```

Before plotting, summarise the data to give the mean value and the 95% confidence intervals ready for plotting.

```{r sumvar-summarise-disparity}
## Summarise disparity data
## Create summary function
myDispRity.Summary <- function(x) {
    na.omit(summary(x, quantile = 95, cent.tend = mean, rounding = 5))
}

## Apply summary to the epoch disparity lists
data_disparitySV_epoch <- lapply(disparitySV_epoch, myDispRity.Summary)
## Apply summary to the bJ disparity lists
data_disparitySV_bJ <- lapply(disparitySV_bJ, myDispRity.Summary)
```

The values of disparity can be seen in the following tables. First for the epoch binning scheme.

```{r sumvar-disparity-tables-epoch, collapse = TRUE, results = "asis"}
## Make table to display disparity values for epoch bins
cat(paste("With negative eigenvalue correction", collapse = "\n"))
pander(data_disparitySV_epoch[1:4])
cat(paste("Without negative eigenvalue correction", collapse = "\n"))
pander(data_disparitySV_epoch[5:8])
```

Then for the 10 Ma bins aligned with the base of the Jurassic.

```{r sumvar-disparity-tables-bJ, collapse = TRUE, results = "asis"}
## For 10 Ma bJ binning
cat(paste("With negative eigenvalue correction", collapse = "\n"))
pander(data_disparitySV_bJ[1:4])
cat(paste("Without negative eigenvalue correction", collapse = "\n"))
pander(data_disparitySV_bJ[5:8])
```

### Rarefaction

Now build some rarefaction curves to show the effect of sample size on the disparity metric. Calculate the disparity on the rarefied subsamples.

```{r sumvar-rarefied-disparity-analyses, cache = TRUE}
## For parallelisation: create cluster
clusDisparity <- makeCluster(no_cores)
## Export required variables
clusterExport(clusDisparity, c("dispRity", "variances"))

## Calculate disparity in epochs
rare_disparitySV_epoch <- parLapply(clusDisparity, rare_bootstrapped_epoch, parDispRitySV)
## Calculate disparity fro bJ binning
rare_disparitySV_bJ <- parLapply(clusDisparity, rare_bootstrapped_bJ, parDispRitySV)

## Release cluster cores
stopCluster(clusDisparity)

## Summarise the data
data_rare_dispSV_epoch <- lapply(rare_disparitySV_epoch, myDispRity.Summary)
data_rare_dispSV_bJ <- lapply(rare_disparitySV_bJ, myDispRity.Summary)
```

Now I plot the rarefaction curves using the inbuilt `plot` function of `dispRity`, which is easier than doing it by hand. <!-- Need to sort this out. -->

```{r sumvar-rarefy-plot, collapse = TRUE, eval = TRUE}
## Plot rarefied data for the epoch bins
# pdf("fig/rarefied-SV-data.pdf", width = 7, height = 7)

## Set up the plot area
# dev.new(width = 7, height = 7)
par(mar = c(1, 1, 3, 1), oma = c(3, 3, 3, 1) + 0.1, mgp = c(1.5, 0.5, 0), lheight = 0.5, bty = "o")

## Get axis labels
plotAxisLabels <- function() {
    mtext("Taxon count", side = 1, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8, font = 2)
    mtext("Sum of variances", side = 2, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8, font = 2)
}

## Get plot names
plotNames <- c("RAW", "GED", "GOW", "MPD", "RAW(uncorrected)", "GED(uncorrected)", "GOW(uncorrected)", "MPD(uncorrected)")

for (dist in 1:length(rare_disparitySV_epoch)) {
    print(dist)
    plot(rare_disparitySV_epoch[[dist]], rarefaction = TRUE, cent.tend = mean, quantile = 95, xlab = "", ylab = "")
    plotAxisLabels()
    mtext(paste0(plotNames[dist], ": epoch-level bins"), side = 3, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 1.0, font = 2)
}
for (dist in 1:length(rare_disparitySV_bJ)) {
    print(dist)
    plot(rare_disparitySV_bJ[[dist]], rarefaction = TRUE, cent.tend = mean, quantile = 95, xlab = "", ylab = "")
    plotAxisLabels()
    mtext(paste0(plotNames[dist], ": 10 Ma bins (base-Jurassic)"), side = 3, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 1.0, font = 2)
}
# dev.off()
```

### Plot disparity by epoch

Now we're ready to plot the disparity data binned by epoch. The functions below find the mean date (midpoint) of each bin to plot the x values, and creates shading boxes to delimit each bin on the plots.

```{r sumvar-disparity-time-epochs-plot-setup}
## Calculate bin means
bin_means_epoch <- rowMeans(disparity_bins$epochs)

## Create function to shade bins in plots
bin.shading.epoch <- function() {
    ## Get bin length for the data
    for (bin in seq(1, nrow(disparity_bins$epochs), 2)) {
        rect(disparity_bins$epochs[bin, "max_age"], 
            par()$usr[4], 
            disparity_bins$epochs[bin, "min_age"], 
            par()$usr[3], 
            density = NA, col = "grey93", xpd = FALSE)
    }
    ## Add an extra bin if the above is even to clarify where the last bin ends
    if (nrow(disparity_bins$epochs) %% 2 == 0) {
        rect(min(disparity_bins$epochs), 
            par()$usr[4], 
            par()$usr[2], 
            par()$usr[3], 
            density = NA, col = "grey93", xpd = FALSE)
    }
}
```

To plot the disparity data, I've created a `for` loop that plots the disparity time series for each distance matrix in a row. Each plot contains two series: the disparity calculated from the ordinated distance matrix with negative eigenvalue correction (red), and without (blue). Legend, axes, and a time scale are plotted at the end.

```{r sumvar-disparity-time-epochs-plot, fig.cap = "Cladistic disparity of 99 ichthyosaur taxa through the Mesozoic. Data is divided into epoch-level bins. Generalised Euclidean (A), Gower (B), and maximum possible distances (C) are shown with data ordinated using principal coordinates analysis with (red) and without (blue) correction for negative eigenvalues (Caillez 1983).", fig.width = 5, fig.height = 12, fig.align = "center", echo = TRUE, warning = FALSE, fig.show = "hold"}
## Set up plot area
# dev.new(width = 5, height = 12)
# cairo_pdf("fig/disparity-time-epoch-sumvar.pdf", width = 5, height = 12)
par(mfrow = c(4, 1), 
	pty = "m", 
	mar = c(0, 0, 0, 0), 
	oma = c(8, 4, 1, 4), 
	mgp = c(1.5, 0.5, 0), 
	lheight = 0.5)

for(nPlot in 1:length(data_disparitySV_epoch[1:4])) {
	nUcPlot <- nPlot + 4
    ## Plot corrected data in red
    plot(bin_means_epoch[1:8], 
        data_disparitySV_epoch[[nPlot]][, "bs.mean"], 
        xlim = rev(range(disparity_bins$epoch)), 
        ylim = range(0, data_disparitySV_epoch[[nPlot]][, "2.5%"], data_disparitySV_epoch[[nPlot]][, "97.5%"]), 
        type = "n", xaxt = "n")
    bin.shading.epoch()
    arrows(bin_means_epoch + 1.5, 
        data_disparitySV_epoch[[nPlot]][, "2.5%"], 
        bin_means_epoch + 1.5, 
        data_disparitySV_epoch[[nPlot]][, "97.5%"], 
        length = 0.02, 
        angle = 90, 
        code = 3, 
        col = "red")
    points(bin_means_epoch + 1.5, 
        data_disparitySV_epoch[[nPlot]][, "bs.mean"], 
        pch = 16, col = "red")
    ## Plot raw uncorrected
    par(new = TRUE)
    plot(bin_means_epoch[1:8], 
        data_disparitySV_epoch[[nUcPlot]][, "bs.mean"], 
        xlim = rev(c(66, 252.17)), 
        ylim = range(0, data_disparitySV_epoch[[nUcPlot]][, "2.5%"], data_disparitySV_epoch[[nUcPlot]][, "97.5%"]), 
        ann = FALSE, axes = FALSE, type = "n")
    arrows(bin_means_epoch - 1.5, 
        data_disparitySV_epoch[[nUcPlot]][, "2.5%"], 
        bin_means_epoch - 1.5, 
        data_disparitySV_epoch[[nUcPlot]][, "97.5%"], 
        length = 0.02, 
        angle = 90, 
        code = 3, 
        col = "blue")
    points(bin_means_epoch - 1.5, 
        data_disparitySV_epoch[[nUcPlot]][, "bs.mean"], 
        pch = 16, col = "blue")
    axis(4)
    box(which = "plot")
}

## Plot time bins and axis
axis_top_epoch <- par()$usr[3]
box_height_epoch <- 0.1 * diff(par()$usr[3:4])
axis_bottom_epoch <- axis_top_epoch - box_height_epoch
axis(1, pos = axis_bottom_epoch)
## Plot time bins
rect(min(par()$usr[1], 252.17), 
    axis_bottom_epoch, 201.3, 
    axis_top_epoch, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(min(par()$usr[1], 252.17), 201.3)), 
    mean(c(axis_bottom_epoch, axis_top_epoch)), 
    "TRIASSIC", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(201.3, 
    axis_bottom_epoch, 
    145, 
    axis_top_epoch, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(201.3, 145)), 
    mean(c(axis_bottom_epoch, axis_top_epoch)), 
    "JURASSIC", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(145, 
    axis_bottom_epoch,
    max(par()$usr[2], 66.0), 
    axis_top_epoch, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(145, max(par()$usr[2], 66.0))), 
    mean(c(axis_bottom_epoch, axis_top_epoch)), 
    "CRETACEOUS", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(par()$usr[1], 
    axis_bottom_epoch, 
    par()$usr[2], 
    axis_top_epoch, 
    col = NA, border = TRUE, xpd = NA)

## Plot axis labels
mtext("Age (Ma)", side = 1, line = 4, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)
mtext("Disparity (sum of variances; negative eigenvalues corrected)", side = 2, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)
mtext("Disparity (sum of variances; negative eigenvalues not corrected)", side = 4, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)

## Add a legend
legend(x = mean(c(par()$usr[1:2])), 
    y = -3 * box_height_epoch, 
    legend = c("Ordinated with negative eigenvalue correction", "Ordinated without negative eigenvalue correction"), 
    xjust = 0.5, 
    yjust = 1, 
    pch = 16, 
    ncol = 1, 
    col = c("red", "blue"),
    bty = "n", 
    xpd = NA, cex = 1.0)

## Stop PDF creation
# dev.off()
```

### Plot disparity the 10 Ma time bins

And then the same with the 10 Ma time bins aligned with the base of the Jurassic.

```{r sumvar-disparity-time-10Ma-bJ-plot, fig.cap = "Cladistic disparity of 99 ichthyosaur taxa through the Mesozoic. Data is divided into 10 Ma bins aligned with the base of the Jurassic (201.3 Ma). Generalised Euclidean (A), Gower (B), and maximum possible distances (C) are shown with data ordinated using principal coordinates analysis with (red) and without (blue) correction for negative eigenvalues (Caillez 1983).", fig.width = 5, fig.height = 12, fig.align = "center", echo = TRUE, warning = FALSE, fig.show = "hold"}
## Get bin mean dates for plotting
bin_means_bJ <- rowMeans(disparity_bins$tenMa_bJ)
bin_means_bJ <- bin_means_bJ[!is.na(summary(disparitySV_bJ[[1]])[, 4])]

## Create function to shade bins in plots
bin.shading.bJ <- function() {
    ## Get bin length for the data
    for (bin in seq(1, nrow(disparity_bins$tenMa_bJ), 2)) {
        rect(disparity_bins$tenMa_bJ[bin, "max_age"], 
            par()$usr[4], 
            disparity_bins$tenMa_bJ[bin, "min_age"], 
            par()$usr[3],
             density = NA, col = "grey93", xpd = FALSE)
    }
    ## Add an extra bin if the above is even to clarify where the last bin ends
    if (nrow(disparity_bins$tenMa_bJ) %% 2 == 0) {
        rect(min(disparity_bins$tenMa_bJ), 
            par()$usr[4], 
            par()$usr[2], 
            par()$usr[3], 
            density = NA, col = "grey93", xpd = FALSE)
    }
}

## Set up plot area
# dev.new(width = 5, height = 12)
# cairo_pdf("fig/disparity-time-bJ-sumvar.pdf", width = 5, height = 12)
par(mfrow = c(4, 1), pty = "m", mar = c(0, 0, 0, 0), oma = c(8, 4, 1, 4), mgp = c(1.5, 0.5, 0), lheight = 0.5)

for(nPlot in 1:length(data_disparitySV_bJ[1:4])) {
	nUCPlot <- nPlot + 4
    ## Plot corrected data in red
    plot(bin_means_bJ, 
        data_disparitySV_bJ[[nPlot]][, "bs.mean"], 
        xlim = rev(range(disparity_bins$tenMa_bJ)), 
        ylim = range(0, data_disparitySV_bJ[[nPlot]][, "2.5%"], data_disparitySV_bJ[[nPlot]][, "97.5%"]), 
        type = "n", 
        xaxt = "n")
    bin.shading.bJ()
    arrows(bin_means_bJ + 1.5, 
        data_disparitySV_bJ[[nPlot]][, "2.5%"], 
        bin_means_bJ + 1.5,
        data_disparitySV_bJ[[nPlot]][, "97.5%"], 
        length = 0.02, 
        angle = 90, 
        code = 3, 
        col = "red")
    points(bin_means_bJ + 1.5, 
        data_disparitySV_bJ[[nPlot]][, "bs.mean"], 
        pch = 16, col = "red")
    ## Plot raw uncorrected
    par(new = TRUE)
    plot(bin_means_bJ, 
        data_disparitySV_bJ[[nUCPlot]][, "bs.mean"], 
        xlim = rev(range(disparity_bins$tenMa_bJ)), 
        ylim = range(0, data_disparitySV_bJ[[nUCPlot]][, "2.5%"], 
            data_disparitySV_bJ[[nUCPlot]][, "97.5%"]), 
        ann = FALSE, axes = FALSE, type = "n")
    arrows(bin_means_bJ - 1.5, 
        data_disparitySV_bJ[[nUCPlot]][, "2.5%"], 
        bin_means_bJ - 1.5, 
        data_disparitySV_bJ[[nUCPlot]][, "97.5%"], 
        length = 0.02, 
        angle = 90, 
        code = 3, 
        col = "blue")
    points(bin_means_bJ - 1.5, 
        data_disparitySV_bJ[[nUCPlot]][, "bs.mean"], 
        pch = 16, col = "blue")
    axis(4)
    box(which = "plot")
}

## Plot time bins and axis
axis_bottom_bJ <- par()$usr[3] - 0.1 * diff(par()$usr[3:4])
axis_top_bJ <- par()$usr[3]
axis(1, pos = axis_bottom_bJ)
## Plot time bins
rect(min(par()$usr[1], 252.17), 
    axis_bottom_bJ, 
    201.3, 
    axis_top_bJ, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(min(par()$usr[1], 252.17), 201.3)), 
    mean(c(axis_bottom_bJ, axis_top_bJ)), 
    "TRIASSIC", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(201.3, 
    axis_bottom_bJ, 
    145, 
    axis_top_bJ, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(201.3, 145)), 
    mean(c(axis_bottom_bJ, axis_top_bJ)), 
    "JURASSIC", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(145, 
    axis_bottom_bJ, 
    max(par()$usr[2], 66.0), 
    axis_top_bJ, 
    col = "white", border = TRUE, xpd = NA)
text(mean(c(145, max(par()$usr[2], 66.0))), 
    mean(c(axis_bottom_bJ, axis_top_bJ)), 
    "CRETACEOUS", 
    adj = c(0.5, 0.5), cex = 1.4, xpd = NA)
rect(par()$usr[1], 
    axis_bottom_bJ, 
    par()$usr[2], 
    axis_top_bJ, 
    col = NA, border = TRUE, xpd = NA)

## Plot axis labels
mtext("Age (Ma)", side = 1, line = 4, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)
mtext("Disparity (mean sum of variances; negative eigenvalues corrected)", side = 2, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)
mtext("Disparity (mean sum of variances; negative eigenvalues not corrected)", side = 4, line = 2, outer = TRUE, adj = c(0.5, 0.5), cex = 0.8)

## Add a legend
legend(x = mean(c(par()$usr[1:2])), y = 2 * axis_bottom_bJ, legend = c("Ordinated with negative eigenvalue correction", "Ordinated without negative eigenvalue correction"), xjust = 0.5, pch = 16, ncol = 1, col = c("red", "blue"), bty = "n", xpd = NA, cex = 1.0)

## Stop PDF creation
# dev.off()
```