# Testing disparity differences

To test between the disparity for each of the above values and binning schemes used, I shall use a series of tests: pairwaise t-tests for differeneces between bins, sequential Wilcox tests for differences between subsequent bins, and Bhattacharrya coefficient for the overlap between bootstrap distributions. All of these will also be subject to correction using the false discovery rate. For for distance matrices, two binning schemes, three disparity metrics, three tests, and two levels of correction, that makes 144 sets of tests. Fun!

For all of these, there will be a print out of the test outcomes to a text file. Variables will be reused as I don't want to have to deal with either numerous variables or very deep lists. Ther are also plenty of `for` loops to make it go smoother.

I also want to check whether differences in the test statitstics are due to the sample sizes, so will repeat with the data rarefied to size of the minimum sample. First, find what that is for each binning scheme. <!-- The rarefied tests need to be added. -->

```{r min-bin-size, }
## For the epoch-bins
min_epoch_bin <- min(size.subsets(subsampled_epoch[[1]]))

## For the 10 Ma bins aligned with the base of the Jurassic
min_bJ_Bin <- min(size.subsets(subsampled_bJ[[1]]))
```

## Pairwise t-tests

First, let's do the pairwise t-tests between the epochs. These test the difference between each pair of time bins, and treats them as independent of each other. There will be a warning for the first in each set as this may increase the possibility of Type I error, but the subsequent FDR correction deals with that.

```{r test-number}
## Number of the test, to separate files
nTestFile <- 1

## Get lists of the required metrics
disp_metric_name <- c("Sum of variances", "Centroid distance", "Diagonal distance")
disp_metric <- c("SV", "Cen", "Dia")
dist_metric <- c("RAW", "GED", "GOW", "MPD", "uRAW", "uGED", "uGOW", "uMPD")
```

I've set this up in loops to do each disparity metric and each distance metric. The main output are tables that show the p values between each pair of bins. Headers are magically added as if by magic.

```{r epochs-ttest, collapse = TRUE, results = "asis"}
for (disp in 1:length(disp_metric)) {
    ## Get the name of the disparity metric
    my_name <- get(paste0("rare_disparity", disp_metric[disp], "_epoch"))

    ## Print a title for this
    cat(paste("###", disp_metric_name[disp], sep = " ", collapse = "\n"))

    ## Start going through distance matrices for t-tests
    for (dist in 1:length(dist_metric)) {
        ## Create list for each distance matrix
        disparity_test <- list()

        ## Do pairwise t-tests
        disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise")))
        ## Repeat with FDR
        disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise", correction = "fdr")))
        ## Repeat with rarefaction
        # disparity_test[[3]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = t.test, comparison = "pairwise", rarefaction = min_epoch_bin, correction = "fdr")))
        # disparity_test[[4]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = t.test, comparison = "pairwise", correction = "fdr", rarefaction = min_epoch_bin)))
        
        ## Add names
        names(disparity_test) <- c("t-test", "t-test with FDR")#, paste("t-test; rarefied to", min_epoch_bin, sep = " "), paste("t-test with FDR; rarefied to", min_epoch_bin, sep = " "))

        ## Split the row names
        for (split in 1:length(disparity_test)) {
            my_Tbl <- do.call(rbind, str_split(rownames(disparity_test[[split]]), " : "))
            disparity_test[[split]]["from"] <- my_Tbl[,1]
            disparity_test[[split]]["to"] <- my_Tbl[,2]
        }

        ## Convert to pairwise table
        nameVals <- sort(unique(unlist(disparity_test[[1]][4:5])))
        disparity_sig_tables <- matrix(0, length(nameVals), length(nameVals), dimnames = list(nameVals, nameVals))

        ## Add values to pairwise table
        disparity_sig_tables[as.matrix(disparity_test[[1]][c("from", "to")])] <- disparity_test[[1]][["p.value"]]
        disparity_sig_tables[as.matrix(disparity_test[[2]][c("to", "from")])] <- disparity_test[[2]][["p.value"]]

        ## Show in document
        cat(paste(dist_metric[dist], sep = " ", collapse = "\n"))
        cat(pander(disparity_sig_tables))

        ## Export test file
        write.csv(disparity_sig_tables, file = paste0("data/", nTestFile, "-", dist_metric[dist], "-", disp_metric[disp], "-ttest.csv"))
        ## Increase the file number
        nTestFile <- nTestFile + 1
    }
}
```

And then for the bJ binning scheme.

```{r bJ-ttest, collapse = TRUE, results = "asis"}
for (disp in 1:length(disp_metric)) {
    ## Get the name of the disparity metric
    my_name <- get(paste0("rare_disparity", disp_metric[disp], "_bJ"))

    ## Print a title for this
    cat(paste("###", disp_metric_name[disp], sep = " ", collapse = "\n\n"))

    ## Start going through distance matrices for t-tests
    for (dist in 1:length(dist_metric)) {
        ## Create list for each distance matrix
        disparity_test <- list()

        ## Do pairwise t-tests
        disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise")))
        ## Repeat with FDR
        disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise", correction = "fdr")))
        ## Repeat with rarefaction
        # disparity_test[[3]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise", rarefaction = min_bJ_bin, correction = "fdr")))
        # disparity_test[[4]] <- as.data.frame(do.call(cbind, test.dispRity(my_name[[dist]], test = t.test, comparison = "pairwise", correction = "fdr", rarefaction = min_epoch_bin)))
        ## Add names
        names(disparity_test) <- c("t-test", "t-test with FDR")#, paste("t-test; rarefied to", min_epoch_bin, sep = " "), paste("t-test with FDR; rarefied to", min_epoch_bin, sep = " "))

 
        ## Split the row names
        for (split in 1:length(disparity_test)) {
            my_Tbl <- do.call(rbind, str_split(rownames(disparity_test[[split]]), " : "))
            disparity_test[[split]]["from"] <- my_Tbl[,1]
            disparity_test[[split]]["to"] <- my_Tbl[,2]
        }

        ## Convert to pairwise table
        nameVals <- sort(unique(unlist(disparity_test[[1]][4:5])))
        disparity_sig_tables <- matrix(0, length(nameVals), length(nameVals), dimnames = list(nameVals, nameVals))

        ## Add values to pairwise table
        disparity_sig_tables[as.matrix(disparity_test[[1]][c("from", "to")])] <- disparity_test[[1]][["p.value"]]
        disparity_sig_tables[as.matrix(disparity_test[[2]][c("to", "from")])] <- disparity_test[[2]][["p.value"]]

        ## Show in document
        cat(paste(dist_metric[dist], sep = " ", collapse = "\n\n"))
        cat(pander(disparity_sig_tables))

        ## Export test file
        # write.csv(disparity_sig_tables, file = paste0("data/", nTestFile, "-bJ-", dist_metric[dist], "-", disp_metric[disp], "-ttest.csv"))
        ## Increase the file number
        nTestFile <- nTestFile + 1
    }
}
```

## Sequential Wilcox tests

The sequential Wilcox tests are used as each time bin is not independent from the previous and each non-parametric.

### Sum of variances

```{r epochs-sumvar-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparitySV_epoch)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Repeat with rarefaction
   # disparity_test[[3]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = wilcox.test, comparison = "sequential", rarefaction = min_epoch_bin, correction = "fdr")))
   # disparity_test[[4]] <- as.data.frame(do.call(cbind, test.dispRity(rare_disparitySV_epoch[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr", rarefaction = min_epoch_bin)))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")#, paste("t-test; rarefied to", min_epoch_bin, sep = " "), paste("t-test with FDR; rarefied to", min_epoch_bin, sep = " "))

    ## Show in document
    cat(paste(plotNames[dist], sep = " ", collapse = "\n\n"))
    # cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-sv-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-sv-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```

And between 10 Ma bins (base-Jurassic).

```{r bJ-sumvar-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparitySV_bJ)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(disparitySV_bJ[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(disparitySV_bJ[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")

    ## Show in document
    cat(paste("####", plotNames[dist], sep = " ", collapse = "\n\n"))
    cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-sv-bJ-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-sv-bJ-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```

### Centroid distance

In epochs.

```{r epochs-centroid-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparityCen_epoch)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(disparityCen_epoch[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(disparityCen_epoch[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")

    ## Show in document
    cat(paste("####", plotNames[dist], sep = " ", collapse = "\n\n"))
    cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-cen-epoch-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-cen-epoch-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```

And for the 10 Ma bins.

```{r bJ-centroid-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparityCen_bJ)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(disparityCen_bJ[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(disparityCen_bJ[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")

    ## Show in document
    cat(paste("####", plotNames[dist], sep = " ", collapse = "\n\n"))
    cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-cen-bJ-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-cen-bJ-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```

### Diagonal distance

For epochs.

```{r epochs-diagonal-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparityDia_epoch)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(disparityDia_epoch[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(disparityDia_epoch[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")

    ## Show in document
    cat(paste(plotNames[dist], sep = " ", collapse = "\n\n"))
    cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-dia-epoch-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-dia-epoch-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```

Between 10 Ma bins.

```{r bJ-diagonal-wilcoxtest, collapse = TRUE, results = "asis"}
## Test disparity differences without and with adjusted p-values (using FDR).
for (dist in 1:length(disparityDia_bJ)) {
    ## Create list for each distance matrix
    disparity_test <- list()

    ## Do sequential Wilcox test
    disparity_test[[1]] <- as.data.frame(do.call(cbind, test.dispRity(disparityDia_bJ[[dist]], test = wilcox.test, comparison = "sequential")))
    ## Repeat with FDR
    disparity_test[[2]] <- as.data.frame(do.call(cbind, test.dispRity(disparityDia_bJ[[dist]], test = wilcox.test, comparison = "sequential", correction = "fdr")))
    ## Add names
    names(disparity_test) <- c("Wilcox test", "Wilcox test with FDR")

    ## Show in document
    cat(paste("####", plotNames[dist], sep = " ", collapse = "\n\n"))
    cat(pander(disparity_test))

    ## Export test file
    # write.csv(disparity_test[[1]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-dia-bJ-swilcox-test.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
    ## Export FDR test file
    # write.csv(disparity_test[[2]], file = paste0("data/", nTestFile, "-", plotNames[dist], "-dia-bJ-swilcox-test-fdr.csv"))
    ## Increase the file number
    nTestFile <- nTestFile + 1
}
```